---
title: K-means exercise in R language
author: Yen-Chung Chen
date: '2018-06-07'
slug: k-means-exercise-in-r-language
categories:
  - R
tags:
  - R
  - k-means
  - statistics
image:
  caption: ''
  focal_point: ''
---
As a novice in genomic data analysis, one of my goal is to benchmark
how well a clustering method works. I ran across [this practice of doing
k-means at R-exercises](https://www.r-exercises.com/2018/04/13/11256/)
the other day and felt it might be a nice start because k-means is easy
to perform and conceptually simple for me to correlate what is happening
behind the clustering machinery.

It starts with manipulating the built-in `iris` dataset as usual. (I
would load `ggplot2` and `cowplot` first.) Now, let’s have some k-means
done.

```{r}
library(ggplot2)
library(dplyr)
library(cowplot)

# Set random seed = 1
set.seed(1)

# Do k-means with 3 centers
iris_k <- kmeans(iris[, c(1,2)], 3)

# Append cluster identity
iris_df <- iris
iris_df$cluster <- iris_k$cluster

# Append cluster identity
iris_df <- iris
iris_df$cluster <- factor(iris_k$cluster)
species <- iris_df$Species
sepal_only <- iris_df$cluster

# Check species proportion in each cluster
print(table(species, sepal_only))

ggplot(iris_df, aes(x = Sepal.Length, y = Sepal.Width, 
                    color = Species, pch = cluster)) +
  geom_point() +
  labs(x = "Sepal Length", y = "Sepal Width", pch = "Cluster") +
  ggtitle("Sepal Only")
```

The cluster identities do not correspond well with species identity, and
the result does not look so impressive. It is imaginable because the
different species intermingle on the sepal dimensions, and `kmeans()`
considering sepal dimensions only would have a hard time in telling them
apart.

```{r}
# Include petal info into k-means
iris_k2 <- kmeans(iris[, c(1:4)], 3)
iris_df$cluster_sp <- factor(iris_k2$cluster)
s_and_p <- iris_df$cluster_sp

# Check the proportion of species of each cluster
print(table(species, s_and_p))

ggplot(iris_df, aes(x = Sepal.Length, y = Sepal.Width, 
                    color = Species, pch = cluster_sp)) +
  geom_point() +
  labs(x = "Sepal Length", y = "Sepal Width", pch = "Cluster") +
  ggtitle("Sepal and Petal")
```

After taking petal length and width, the clustering identities agreed
more with the “real” species identities.

## Effect of linear transformation: What would happen if we multiply `Petal.Width` by 2?


```{r}
# Multiply Petal.width by 2 and do k-means again
iris_df$Petal.Width <- iris_df$Petal.Width * 2
iris_k_2pw <- kmeans(iris_df[, c(1:4)], 3)
iris_df$cluster_pw2 <- factor(iris_k_2pw$cluster)
doubled <- iris_df$cluster_pw2

# Estimate how much cluster identity agrees with each other before
# and after doubling petal width
print(table(s_and_p, doubled))

ggplot(iris_df, aes(x = Sepal.Length, y = Sepal.Width, 
                    color = Species, pch = cluster_pw2)) +
  geom_point() +
  labs(x = "Sepal Length", y = "Sepal Width", pch = "Cluster") +
  ggtitle("2 x Petal Width")
```

Though the clustering result did not change too much, linear
transformation indeed altered the clustering identity. I was not
expecting that at first, but then *I realized k-means takes only
distance into consideration, and linear transformation on one dimension
does change relative distance.* So, while transforming one dimension
would influence clustering result, multiplying everything at the same
time should not change cluster identity.

```{r}
# Multiply everything by 2 and do k-means again
iris_df2 <- iris_df * 2
iris_k_db <- kmeans(iris_df2[, c(1:4)], 3)

all_doubled <- iris_k_2pw$cluster

# Estimate how much cluster identity agrees with each other before
# and after doubling petal width
print(table(all_doubled, doubled))

```


And that’s definitely the case.

## Scaling data

One way to think of it might be that k-means considers Euclidean
distance, and if we expand or shrink one dimension, the influence of
that dimension on distance would change accordingly. To mitigate this
asymmetry of influence, scaling might be a good way if we assume every
dimension should have equal impact in clustering.

```{r}
# Doing z-transformation with scale()
iris_df[, c(1:4)] <- scale(iris_df[, c(1:4)])
iris_k_z <- kmeans(iris_df[, c(1:4)], 3)

iris_df$cluster_kz <- iris_k_z$cluster
z_trans <- iris_df$cluster_kz

# Doing z-transformation with scale() of the original iris
iris_ori <- scale(iris[, c(1:4)])
iris_k_zo <- kmeans(iris_ori[, c(1:4)], 3)
z_before_dbl <- iris_k_zo$cluster

# Before and after doubling with z-transformation
print(table(z_before_dbl, z_trans))
```

Scaling do a good job to give us consistent result regardless of whether
there is linear transformation. It is thus advisable to scale the data
in some way to make sure we could capture the diversity fairly and would
not let the dimension largest in number dominate the whole clustering.

## Yet another dataset

Now, we move on to another dataset from
[Kaggle.com](https://www.kaggle.com/c/titanic/data). The author of this
exercise suggested us to cluster passengers according to *sex, number of
sibling or spouse on board (SibSp), number of children or parents on
board (Parch), and ticket fare (fare)*, and see if the cluster could
predict survival.

```{r}
# Playing with training dataset from Titanic@Kaggle
csvurl <- "https://storage.googleapis.com/kaggle-competitions-data/kaggle/3136/train.csv?GoogleAccessId=web-data@kaggle-161607.iam.gserviceaccount.com&Expires=1553899686&Signature=eUG7kTwcd4FwTAPYZeKwgJafj6bF77hbvk8Exk%2BgM99szovTUfGsMh7%2F6zoYLlnVQ%2Fmf%2BNRHWKNcpyJpUDXBdf3Vr5UpETVUiOn5i90HV2%2FNfLeQ4uc%2Bap1KfmcyfHEojrDViahLWKrNUTaLvGrlrItZhoo9UWFyf6hn3n%2BZtC4Iior1S0mDk%2FrXb51sQJyzpIMy2d%2Fgzqtt7duZJcOxaX7EvHvefjGSc7ZJAKDumqGLxXKGaA4lUVa76ELsC53cqxLDC8oSHCMpj0afDcCUvSdbmCgvImWyINKKAWupHEmmNrKRDN%2FsL%2F0vS2MTvbEWC8%2B8C98lsSUlkxwxpqqFxA%3D%3D"

train <- read.csv(csvurl, stringsAsFactors = F)
print(head(train))

# Make Sex and Pclass dummy variable
train$Sex <- factor(train$Sex, levels = c("female", "male"),
                    labels = c("0", "1"))
train$Pclass <- ifelse(train$Pclass == "3", 1, 0)

# K-means (k = 4) according to Sex, SibSp, Parch, Fare
tita_k <- kmeans(select(train, Sex, SibSp, Parch, Fare), centers = 4, nstart = 20)
train$cluster_1 <- tita_k$cluster

# Survival of each cluster
print(
  table(train$Survived, train$cluster_1)[2, ]/colSums(table(train$Survived, train$cluster_1))
  )
```

It does look like the rate of survival differs between groups. To
evaluate how big k should be to capture the most information from the
dataset, variance explained could be calculated here. It was actually
super-easy because `kmeans()` already prepared us *the sum of squares
between clusters* (betweenss) and *total sum of squares* (totss) in its
output.

```{r}
# Trying different k (2 - 20)
## Initiate a list to store the results
tita_km <- list()

for (i in c(2:20)) {
  index <- i - 1
  tita_km[[index]] <- kmeans(select(train, Sex, SibSp, Parch, Fare), centers = i, nstart = 20)
}

# Calculate variance explained by cluster
ve <- sapply(tita_km, function(x) x[["betweenss"]]/x[["totss"]])
plot(ve ~ c(2:20), xlab = "k",
     ylab = "SoS between clusters / Total SoS",
     col = c(rep("black",3), "red", rep("black",15)), pch = 16)
```

With visualization, it seems the margin drops when k \> 5, and thus [the
elbow
method](https://en.wikipedia.org/wiki/Elbow_method_%28clustering%29)
would suggest k = 5 as the optimal cluster number.

```{r}
train$cluster_5 <- tita_km[[5]][["cluster"]]
print(
  table(train$Survived, train$cluster_5)[2, ] /colSums(table(train$Survived, train$cluster_5))
)

```

This time, the difference of survival between clusters becomes even
sharper. Does more variance explained necessarily lead to better
correlation with outcome of interest? To test this notion, I went back
to do the same thing for the iris dataset.

```{r}
# Testing different k in iris

iris_km <- list()

for (i in c(2:20)) {
  index <- i - 1
  iris_km[[index]] <- kmeans(select(iris, -Species), centers = i, nstart = 20)
}

# Calculate variance explained
ve_iris <- sapply(iris_km, function(x) x[["betweenss"]]/x[["totss"]])
plot(ve_iris ~ c(2:20), xlab = "k",
     ylab = "SoS between clusters / Total SoS",
     col = c(rep("black",2), "red", rep("black",16)), pch = 16)
```

It seems that the “elbow” is on k = 4.

```{r}
# k = 4
iris$clus4 <- iris_km[[3]][["cluster"]]
k4 <- as.data.frame(table(iris$Species, iris$clus4))

k4p <- ggplot(k4, aes(x = Var2, y = Freq, fill = Var1)) +
  geom_bar(stat = "identity") +
  labs(x = "Cluster", fill = "Species") +
  ggtitle("k = 4 for iris")

# k = 3
iris$clus3 <- iris_km[[2]][["cluster"]]
k3 <- as.data.frame(table(iris$Species, iris$clus3))

k3p <- ggplot(k3, aes(x = Var2, y = Freq, fill = Var1)) +
  geom_bar(stat = "identity") +
  labs(x = "Cluster", fill = "Species") +
  ggtitle("k = 3 for iris")

print(plot_grid(k4p, k3p, labels = "AUTO"))
```

It seems that versicolor and virginica are hard to tell apart based on
the information iris dataset provides, but k = 4 did a better job in
identifying a subset of versicolor apart from the mixed population. In
this case, k number indicated by elbow method seems to score again.

## Take home message

This exercise helped me learn the fundamental behaviors of k-means and
how R implemented it among other things. It was a pleasant surprise to
discover `scale()` , so I no longer have to reinvent the wheel. The
detailed results in addition to cluster identity `kmeans()` provides by
default not only make life easier and remind me of the things to look
after clustering.

1.  K-means sort of *summarizes* information of multiple
    dimension to categorize them.
2.  K-means is distance-based, so linear transformation
    would change the clustering result.
3.  To prevent the dimensions bigger in number dominate
    clustering, scaling is a recommended before k-means if we assume
    every dimension is equal in importance.
4.  Besides cluster identity, `kmeans()` also gives
    total sum of squares, sum of square between clusters, and in-cluster
    sum of squares, and many other information by default
    (check `?kmeans()`). This makes calculation of variance explained
    very intuitive.
5.  Elbow method is based on the marginal gain of
    variance explained with adding more and more cluster and could help
    us assess the preferable number of cluster.
